An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-07-30 EconGrowth2.Rnw
New file for the actual paper.

\section*{To-do List}
\begin{itemize}
\item{Heun: Figure out why CES$q$ model fails to converge with original data for JP and CN.}
\item{Heun: Investigate why wew obtain $\rho_1 < -1$ 
for some US CES$e$ resample models.}
\item{Pruim: Respond to Heun's wish list for graphs.}
\item{Heun: Figure out why CESe for ZA with $x$ fails during resampling.}
\item{Heun: Review all assumptions for CES with energy models. 
There are several assumptions for rho and rho1 for various countries. 
When we are resampling, we probably don't want to specify those values.}
\item{Heun: Investigate possible conflicting use of exergy by Warr and Ayres.}
\item{Heun: Fix the energy sankey diagram from the seminar}
\item{Heun: Include adjusted sankey diagram in the seminar}
\item{Heun: Add CES production function fits}
      \begin{itemize}
      \item{Write up the decisions about parameter estimation for the 
            CES production function fits}
      \item{Revisit my assumptions for the CES production function 
            (with energy) for each country.}
      \item{Fix the problems with $\sigma_1$ and $\sigma$ values in 
            the CES parameter tables in the appendix}
      \item{Add CES rows to the AIC table. Code is in the 
            \texttt{createAICTable} function, but it is not yet working. 
            Need to find a way to calculate AIC for the object that is 
            returned from the \texttt{cesEst} function.}
      \end{itemize}
\item{Heun: Fix warnings of the form \texttt{Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562} in the code that generates the Cobb-Douglas with energy fits.}
\item{de Wit: review, critique, and contextualize.}
\item{\st{Heun: Add option for $m = 1$ in single-factor models.}}
\item{\st{Heun: Re-implement confidence interval calculations to use
resampling methods.}}
      \begin{itemize}
      \item{\st{SF}}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{\st{CES}}
      \item{\st{CESe}}
      \item{\st{LINEX}}
      \end{itemize}
\item{\st{Heun: Include scatter graphs for model parameters 
estimated by resampling methods.}}
      \begin{itemize}
      \item{\st{SF}}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{\st{CES}}
      \item{\st{CESe}}
      \item{\st{LINEX}}
      \end{itemize}
\item{\st{Heun: Decide if the single factor models should include constraints
on $\alpha$, $\beta$, and $\gamma$. At present, the single-factor model allows
values outside the ranges $\alpha \in (0,1)$, $\beta \in (0,1)$, and $\gamma \in (0,1)$.}}
\item{\st{Heun: Make smaller scatter plots for resample coefficients for those 
countries where $u$ is involved.}}
\item{\st{Heun: Add a table that shows the fraction of unconverged resample fits
for each model, energy type, and factor.}}
\item{\st{Heun: Get AIC calculations working again.}}
\item{\st{Pruim: Add base fit red dots to all cloud plots.}}
      \begin{itemize}
      \item{\st{SF}}
      \item{\st{CD}}
      \item{\st{CDe}}
      \item{\st{CES}}
      \item{\st{CESe}}
      \item{\st{LINEX}}
      \end{itemize}
\item{\st{Heun: Look into constraining $\delta \in [0,1]$
and $\delta_1 \in [0,1]$ in the CES and CES$e$ models.}}
\end{itemize}

Neoclassical economic growth theory attributes output ($Y$) to factors of production:
capital stock ($K$) and labor ($L$). The Cobb-Douglas production function is the 
canonical representation of indexed economic output ($y$) as a function of indexed 
labor ($l$), indexed capital stock ($k$), and an exogenous, time-dependent 
technological progress term ($A$):

\begin{equation} \label{eq:CD}
  y = A k^\alpha l^\beta; A = \mathrm{e}^{\lambda(t-t_0)}
\end{equation}

\noindent where $y \equiv Y/Y_\mathrm{0}$, $k \equiv K/K_\mathrm{0}$, and 
$l \equiv L/L_\mathrm{0}$; with $Y$ represented by GDP, $K$ expressed in 
currency units, $L$ expressed in workers or work-hours/year, time ($t$) measured in years, 
with e being the base of the natural logarithm, and subscripts ``0'' indicating values 
at an initial year. $A$ is a unitless technological progress term, with $\lambda$ 
indicating the pace of technological progress (in units of 
year$^{-1}$).\footnote{Dimensionless, indexed quantities are represented by 
lower-case symbols 
($y$, $k$, $l$, $e$, $q$, $x$, and $u$), and dimensional 
quantities are represented by upper-case symbols 
($Y$, $K$, $L$, $E$, $Q$, $X$, and $U$). One exception is the technolgical 
progress term $A$, which is dimensionless but written upper-case by convention.} 
Neoclassical economic growth theory as expressed in the Cobb-Douglas production 
function makes no attempt to describe the mechanisms by which technological 
progress ($A$) affects economic growth, nor does it attempt to describe how 
technological progress ($A$) might vary from one economy to the next. Mathematically, 
$A$ is a fitting parameter that aligns the predictions of Equation~\ref{eq:CD} 
with historical data. Typically, the constraint $\alpha + \beta = 1$ is assumed, 
indicating constant returns to scale.

As shown in Equation~\ref{eq:CD} above, neoclassical economic theory does not 
include an explicit term for the role of energy in economic growth. However, 
\citet{Bashmakov:2007ek}, \citet{Hamilton:2009ui}, \citet{Murphy:2011jh}, 
\citet{Benes:2012tm}, \citet{Kumhof:2012ua}, and others indicate that the 
importance of energy for economic growth may far outweigh its cost share in the 
economy. That neoclassical economic growth theory ignores energy is, therefore, 
cause for concern, and inquiry into the proper approach for including energy 
as a factor of production in economic growth modeling is both relevant and important.

Some energy economists augment economic growth models by including energy as an 
additional (or the only) factor of production. However, there is little agreement 
on how this should be done. Neither the mathematical form of the energy-augmented 
production function nor the quantification of energy are consistent in the 
literature. There has been little empirical work assessing the predictive ability 
of various energy-augmented production functions for different types of economies. 
And, very few studies of the importance of energy in economic growth include 
rigorous statistical assessments of energy-related fitting parameters. This paper 
attempts to contribute to the literature by applying statistically rigorous 
methods along three coordinates of analysis: form of the energy augmented 
production function, energy quantification, and type of economy.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parameter Estimation}
\label{sec:Parameter_Estimation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Each of the economic growth models discussed in 
Section~\ref{sec:Prod_Func_Forms} has parameters which must be estimated 
for each economy (see Section~\ref{sec:Economy_Types}) and 
each energy quantification (see Section~\ref{sec:Quantification_of_Energy}) 
via least-squares methods. The following subsections discuss 
parameter estimation methods and interpret the results.

%++++++++++++++++++++++++++++++
\subsection{Cobb-Douglas Models}
\label{sec:CD_Parameter_Estimation}
%++++++++++++++++++++++++++++++

The Cobb-Douglas model without energy is given by 
Equation~\ref{eq:CD}. The Cobb-Douglas production function was
reparameterized as

\begin{equation} \label{eq:CD_reparam}
  y = A k^{\alpha} l^{1-\alpha}; A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

\noindent to ensure $\alpha + \beta = 1$ for constant returns to scale. 
$\lambda$ and $\alpha$ were estimated by the 
\texttt{R} \citep{R} function \texttt{nls}. 
If the estimated value for $\alpha$ was found outside the interval $[0,1]$, 
we set $\alpha$ to its boundary value and re-estimated $\lambda$. 
The value of $\beta$ was found with $\beta = 1 - \alpha$.

Figure~\ref{fig:CD_Cloud_Graph} shows $\alpha$, $\beta$, and $\gamma$ parameter values for resample models for the Cobb-Douglas model (without energy). $\gamma$ is always 0.0, because energy is not included in this model.

<<CD_Cloud_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=ninePanelTernaryGraphHeight, fig.cap="Cobb-Douglas (without energy) resample model parameters.">>=
cdResampleTrianglePlot(energyType=NA)
@

Converting the resample data into 95\% confidence intervals 
provides data for Figure~\ref{fig:CD_Params_Graph}.
Appendix~\ref{sec:Model_Parameters} 
contains tables with numerical values of 
Cobb-Douglas model parameters and 95\% confidence intervals.

<<CD_Params_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=cdParameterGraphHeight, fig.cap="Cobb-Douglas (without energy) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph(energyType=NA)
@

To qualitatively assess whether augmenting the Cobb-Douglas model with energy 
will be beneficial, we created partial residual plots as shown in 
Figures~\ref{fig:CDPartialResidualsQ}--\ref{fig:CDPartialResidualsU} below. 
If a monotonic trend is observed on the graphs, we expect that 
energy augmentation will be beneficial. 
However, no strong monotonic tends are observed for any country, 
with the exception, perhaps, of the United Kingdom, whose residuals are, 
on average, above 0.

<<CDPartialResidualsQ, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.align="center", fig.cap="Cobb-Douglas partial residuals plot for thermal energy ($q$).">>=
createPartialResidualPlot(modelType="CD", energyType="Q", countryAbbrev=NA)
@

<<CDPartialResidualsX, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.align="center", fig.cap="Cobb-Douglas partial residuals plot for exergy ($x$).">>=
createPartialResidualPlot(modelType="CD", energyType="X", countryAbbrev=NA)
@

<<CDPartialResidualsU, fig.pos="H", fig.width=threePanelGraphWidth, fig.height=threePanelGraphHeight, fig.align="center", fig.cap="Cobb-Douglas partial residuals plot for useful work ($u$).">>=
createPartialResidualPlot(modelType="CD", energyType="U", countryAbbrev=NA)
@

To estimate the parameters $\lambda$, $\alpha$, $\beta$, and $\gamma$ 
in the energy-augmented Cobb-Douglas model, Equation~\ref{eq:CDe} 
was reparameterized as

\begin{equation} \label{eq:CDe_AB_reparam}
  y = A k^{\min(a,b)} l^{|b-a|} e^{1 - \max(a,b)}; A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

\noindent with 
$\alpha = \min(a,b)$, 
$\beta = |b-a|$, and 
$\gamma = 1 - \max(a,b)$.
The reparameterization ensures that $\alpha + \beta + \gamma = 1$. 
The \texttt{R} \citep{R} function \texttt{nls} was used to estimate 
values of $\lambda$, $a$, and $b$.
If the fitted value for $a$ ($b$) fell outside $[0,1]$, it was 
set equal to its boundary value and $\lambda$ and $b$ ($a$) 
were re-estimated.

To demonstrate the characteristics of the parameter estimation process, 
we performed a grid search over values of $\gamma$, 
wherein we re-estimated values of $\lambda$, $\alpha$, and 
$\beta$ at each value of $\gamma$. 
Figure~\ref{fig:CDeGridLattice} shows the sum of squared errors ($SSE$), 
$\alpha$, and $\beta$ as a function of the fixed value of 
$\gamma$ with thermal energy ($q$).

**** Fix the right y axis label. SSE should be italicized ($SSE$). 
Also, scaling on the axis labels and scales is not being picked up. ****

<<CDeGridLattice, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="Grid search over $\\gamma$ for Cobb-Douglas model with thermal energy ($q$).">>=
cdeGridLattice(energyType="Q")
@

Figure~\ref{fig:CDeGridLattice} shows that, for most economies,
the minimum $SSE$ occurs at or near a boundary of 
$\alpha$, $\beta$, or $\gamma$. 
The parameter estimation algorithm (\texttt{nls}) had difficulty converging when
the optimal value for the parameter represented by the absolute value function 
(in Equation~\ref{eq:CDe_AB_reparam}, $\beta$) 
was near a boundary. If the absolute value caused non-convergence of the 
\texttt{nls} function, second and, if necessary, third reparameterizations 
were attempted:

\begin{equation} \label{eq:CDe_CD_reparam}
  y = A k^{1 - \max(c,d)} l^{\min(c,d)} e^{|d-c|}; A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

\begin{equation} \label{eq:CDe_FG_reparam}
  y = A k^{|g-f|} l^{1 - \max(f,g)} e^{\min(f,g)}; A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

\noindent If none of the three reparameterizations converged, 
the parameterization with the smallest $SSE$
was used. \ref{sec:Convergence_Resample_Models} shows the 
fraction of unconverged models for each model type. The above approach
resulted in very few unconverged resample models, 
even with \Sexpr{numResamples()} resamples.

Figure~\ref{fig:CDq_Cloud_Graph} shows $\alpha$, $\beta$, and $\gamma$ 
parameter values for Cobb-Douglas (with $q$) resample models.

<<CDq_Cloud_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=ninePanelTernaryGraphHeight, fig.cap="Cobb-Douglas (with $q$) resample model parameters.">>=
cdResampleTrianglePlot(energyType="Q")
@

Figure~\ref{fig:CDx_Cloud_Graph} shows $\alpha$, $\beta$, and $\gamma$ parameter values for resample models for the Cobb-Douglas model (with $x$).

<<CDx_Cloud_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=ninePanelTernaryGraphHeight, fig.cap="Cobb-Douglas (with $x$) resample model parameters.">>=
cdResampleTrianglePlot(energyType="X")
@

Figure~\ref{fig:CDu_Cloud_Graph} shows $\alpha$, $\beta$, and $\gamma$ parameter values for resample models for the Cobb-Douglas model (with $u$).

<<CDu_Cloud_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=threePanelTernaryGraphHeight, fig.cap="Cobb-Douglas (with $u$) resample model parameters.">>=
cdResampleTrianglePlot(energyType="U")
@

Figure~\ref{fig:CDq_Params_Graph} shows values and 95\% confidence intervals 
for the parameters for the Cobb-Douglas model (with $q$).

<<CDq_Params_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=cdParameterGraphHeight, fig.cap="Cobb-Douglas (with $q$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph(energyType="Q")
@

The Cobb-Douglas (with exergy) parameters are given in 
Table~\ref{tab:CD_Parameters_With_X}.

Figure~\ref{fig:CDx_Params_Graph} shows values and 95\% confidence intervals 
for the parameters for the Cobb-Douglas model (with $x$).

<<CDx_Params_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=cdParameterGraphHeight, fig.cap="Cobb-Douglas (with $x$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph(energyType="X")
@

Figure~\ref{fig:CDu_Params_Graph} shows values and 95\% confidence intervals 
for the parameters for the Cobb-Douglas model (with $u$).

<<CDu_Params_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=cdParameterGraphHeight, fig.cap="Cobb-Douglas (with $u$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph(energyType="U")
@

Figure~\ref{fig:CD_GDP_Lattice_Graph} compares predictions 
from the Cobb-Douglas models 
(without energy, with $e=q$, with $e=x$, and with $e=u$) to historical data.

<<CD_GDP_Lattice_Graph, fig.pos="H", fig.width=cdParameterGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="Cobb-Douglas results.">>=
createCDLatticeGraph()
@

%++++++++++++++++++++++++++++++
\subsection{CES Models}
\label{sec:CES_Parameter_Estimation}
%++++++++++++++++++++++++++++++

% This bit about the Stern model is commented out for now. I'm leaving it in the paper in case we decide to use it later.
% \citet{Stern:2012ey} developed a CES model of the form\footnote{The CES equation in \citet{Stern:2012ey} appears as
% 
% \begin{equation*} \label{eq:Stern_CES_model}
%   y = \left\{ (1-\gamma_E)^{1/\sigma} \left[ k^{\alpha} \left( \mathrm{e}^{\lambda_L(t-t_0)} l \right)^{\beta} \right]^{\frac{\sigma-1}{\sigma}}   
%       + (\gamma_E)^{1/\sigma} \left[ \mathrm{e}^{\lambda_E(t-t_0)} e \right]^{\frac{\sigma - 1}{\sigma}}   \right\}
%       ^{\frac{\sigma}{\sigma - 1}},
% \end{equation*}
% 
% but the $1/\sigma$ exponent on the $(1-\gamma_E)$ and $\gamma_E$ terms is erroneous due to $y_{\mathrm{t_0}} \neq 1$.} 
% 
% 
% \begin{equation} \label{eq:Stern_CES_model_in_print}
%   y = \left\{ (1-\gamma_E) \left[ k^{\alpha} \left( \mathrm{e}^{\lambda_L(t-t_0)} l \right)^{\beta} \right]^{\frac{\sigma-1}{\sigma}}   
%       + (\gamma_E) \left[ \mathrm{e}^{\lambda_E(t-t_0)} e \right]^{\frac{\sigma - 1}{\sigma}}   \right\}
%       ^{\frac{\sigma}{\sigma - 1}},
% \end{equation}
% 
% \noindent where $e$ is one of $q$, $x$, or $u$; $\sigma$, $\gamma_E$, $\beta$, $\lambda_L$, and $\lambda_E$ are fitting parameters; $\mathrm{e}$ is the base of the natural logarithm; and the sum $\alpha + \beta$ is constrained to unity.

The CES model without and with energy is given by 
Equations~\ref{eq:CES} and \ref{eq:CESe}, respectively. 
The \texttt{R} \citep{R} package \texttt{micEconCES} 
\citep{Henningsen:2011td} was used to estimate parameters 
$\gamma$, $\lambda$, $\delta_1$, $\rho_1$, $\delta$, and $\rho$. 
**** More details here about which algorithms were used, etc. 
Complete after our work this summer. ****

% \citet{Henningsen:2011td}, in their detailed analysis of 
% \citet{Kemfert:1998vd}, found that 
% 
% \begin{quote}
% \ldots the Levenberg-Marquardt and the PORT algorithms 
% are---at least in this study---most likely to find the 
% coefficients that give the best fit to the model, 
% where the PORT algorithm can be used to restrict the estimates 
% to the economically meaningful region.
% \end{quote}

Coefficients for each resampled CES model (without energy)
are shown in 
Figures~\ref{fig:CES_lambda_gamma_Cloud_Graph}--\ref{fig:CES_sigma_delta_Cloud_Graph}.

<<CES_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesResamplePlotLambdaGamma(energyType=NA)
@

<<CES_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesResamplePlotSigma_1Delta_1(energyType=NA)
@

The next graph is not particularly interesting, 
because $\delta \rightarrow 1.0$, 
$\rho \rightarrow -1$, and 
$\sigma \rightarrow \infty$
for the CES without energy model.

<<CES_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (without energy) resample model parameters $\\rho$ and $\\delta$.">>=
cesResamplePlotSigmaDelta(energyType=NA)
@

Figure~\ref{fig:CES_Params_Graph} shows values and 95\% confidence intervals 
for the parameters for the CES model (without energy).

<<CES_Params_Graph, fig.pos="H", fig.width=cesParameterGraphWidth, fig.height=cesParameterGraphHeight, fig.cap="CES (without energy) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCESParamsGraph(energyType=NA)
@

Similar to the analysis of the Cobb-Douglas model in 
Section~\ref{sec:CD_Parameter_Estimation}, 
we make partial residual plots for the CES model (without energy) in 
Figures~\ref{fig:CESPartialResidualsQ}--\ref{fig:CESPartialResidualsU} below. 
Again, no strong monotonic trend is observed, suggesting the 
energy-augmentation of the CES model will not significantly improve the model.

<<CESPartialResidualsQ, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.align="center", fig.cap="CES partial residuals plot for thermal energy ($q$).">>=
createPartialResidualPlot(modelType="CES", energyType="Q", countryAbbrev=NA)
@

<<CESPartialResidualsX, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.align="center", fig.cap="CES partial residuals plot for exergy ($x$).">>=
createPartialResidualPlot(modelType="CES", energyType="X", countryAbbrev=NA)
@

<<CESPartialResidualsU, fig.pos="H", fig.width=threePanelGraphWidth, fig.height=threePanelGraphHeight, fig.align="center", fig.cap="CES partial residuals plot for useful work ($u$).">>=
createPartialResidualPlot(modelType="CES", energyType="U", countryAbbrev=NA)
@

Coefficients for each resampled CES model (with $q$)
are shown in 
Figures~\ref{fig:CESq_lambda_gamma_Cloud_Graph}--\ref{fig:CESq_sigma_delta_Cloud_Graph}.

<<CESq_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesResamplePlotLambdaGamma(energyType="Q")
@

<<CESq_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesResamplePlotSigma_1Delta_1(energyType="Q")
@

<<CESq_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $q$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesResamplePlotSigmaDelta(energyType="Q")
@

Coefficients for each resampled CES model (with $x$)
are shown in 
Figures~\ref{fig:CESx_lambda_gamma_Cloud_Graph}--\ref{fig:CESx_sigma_delta_Cloud_Graph}.

<<CESx_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesResamplePlotLambdaGamma(energyType="X")
@

<<CESx_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesResamplePlotSigma_1Delta_1(energyType="X")
@

<<CESx_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES (with $x$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesResamplePlotSigmaDelta(energyType="X")
@

Coefficients for each resampled CES model (with $u$)
are shown in 
Figures~\ref{fig:CESu_lambda_gamma_Cloud_Graph}--\ref{fig:CESu_sigma_delta_Cloud_Graph}.

<<CESu_lambda_gamma_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\lambda$ and $\\gamma$.">>=
cesResamplePlotLambdaGamma(energyType="U")
@

<<CESu_sigma_1_delta_1_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\sigma_1$ and $\\delta_1$.">>=
cesResamplePlotSigma_1Delta_1(energyType="U")
@

<<CESu_sigma_delta_Cloud_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="CES (with $u$) resample model parameters $\\sigma$ and $\\delta$.">>=
cesResamplePlotSigmaDelta(energyType="U")
@

Figure~\ref{fig:CESq_Params_Graph} shows values and 95\% confidence intervals 
for the parameters for the CES model (with $q$).

<<CESq_Params_Graph, fig.pos="H", fig.width=cesParameterGraphWidth, fig.height=cesParameterGraphHeight, fig.cap="CES (with $q$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCESParamsGraph(energyType="Q")
@

Figure~\ref{fig:CESx_Params_Graph} shows values and 95\% confidence intervals 
for the parameters for the CES model (with $x$).

<<CESx_Params_Graph, fig.pos="H", fig.width=cesParameterGraphWidth, fig.height=cesParameterGraphHeight, fig.cap="CES (with $x$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCESParamsGraph(energyType="X")
@

Figure~\ref{fig:CESu_Params_Graph} shows values and 95\% confidence intervals 
for the parameters for the CES model (with $u$).

<<CESu_Params_Graph, fig.pos="H", fig.width=cesParameterGraphWidth, fig.height=cesParameterGraphHeight, fig.cap="CES (with $u$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCESParamsGraph(energyType="U")
@

Figure~\ref{fig:CES_GDP_Lattice_Graph} compares CES model predictions 
to historical data.

<<CES_GDP_Lattice_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="CES results.">>=
createCESLatticeGraph()
@

%++++++++++++++++++++++++++++++
\subsection{Confidence Intervals for Cobb-Douglas and CES Energy Parameters}
\label{sec:CI_for_CD_and_CES}
%++++++++++++++++++++++++++++++

The key energy parameters for the Cobb-Douglas and CES models are 
$\gamma$, $\delta$, and $\sigma$. 
The parameter $\gamma$ in the Cobb-Douglas model (Equation~\ref{eq:CDe}) 
indicates the importance of energy for economic growth. 
When $\gamma \rightarrow 0$, energy is not a significant contributor 
to economic growth for an economy. 
Conversely, when $\gamma \rightarrow 1$, energy is the primary contributor 
to economic growth. 
However, if the 95\% confidence interval for $\gamma$ for any given economy 
extends to 0.0, we cannot say with high confidence whether or not energy 
makes a meaningful contribution to economic growth for that economy. 
On a graph of the width of the 95\% confidence interval for 
$\gamma$ ($CI_{95\%,\gamma}$) vs. $\gamma$, where each economy is represented 
by a point (its two-letter abbreviation; see Table~\ref{tab:Economy_Examples}), 
the locus of points where the 95\% confidence interval for 
$\gamma$ crosses 0 is a diagonal line. 
If the combination of $\gamma$ and its confidence interval 
for a given economy is below and to the right of the line, 
we have high confidence that we can say whether or not energy 
is a significant contributor to economic growth. 
Thus, only when $\gamma$ is below the line and small (i.e., close to 0) 
for an economy, can we say with confidence that energy is 
an insignificant contributor to economic growth. 
Only when $\gamma$ is below the line and large (i.e., close to 1) 
for an economy can we say that energy is a significant contributor 
to economic growth, according to the Cobb-Douglas model.

The left column of graphs in Figure~\ref{fig:CIvsParam_Lattice_Graph} shows 
$CI_{95\%,\gamma}$ vs. $\gamma$ for the three quantifications of energy 
(thermal energy, $q$, exergy, $x$, and useful work, $u$). 
We note that most economies are above and to the left of the line, 
while only a few are below and to the right.

For the CES model (Equation~\ref{eq:CESe}), the parameter $\delta$ 
indicates the importance of energy for economic growth. 
When $\delta \rightarrow 1$, energy is an insignificant contributor 
to economic growth, and when $\delta \rightarrow 0$, 
energy is the primary contributor to economic growth for an economy. 
However, if the 95\% confidence interval for $\delta$ extends to 1, 
we cannot say with high confidence whether or not energy makes 
a meaningful contribution to economic growth for that economy. 
In a manner similar to $\gamma$ for the Cobb-Douglas model, 
we note that if the combination of $\delta$ and its confidence interval 
for a given economy is below and to the left of the line, 
we have high confidence that we can say whether or not energy is 
a significant contributor to economic growth. 
Thus, only when $\delta$ is below the line and large (i.e., close to 1) 
for an economy, can we say with confidence that energy is 
an insignificant contributor to economic growth. 
Only when $\delta$ is below the line and small (i.e., close to 0) 
for an economy can we say that energy is a significant contributor 
to economic growth, according to the CES model.

The middle column of graphs in Figure~\ref{fig:CIvsParam_Lattice_Graph} shows 
$CI_{95\%,\delta}$ vs. $\delta$ for the three quantifications of energy 
(thermal energy, $q$, exergy, $x$, and useful work, $u$). 
We note that most economies are above and to the right of the line, 
while only a few are below and to the left.

For the CES model (Equation~\ref{eq:CESe}), the parameter $\sigma$ indicates 
whether capital stock and labor ($kl$) and energy ($e$) are 
complements ($\sigma <$ 1) or substitutes ($\sigma >$ 1). 
For countries whose combination of  $\sigma$ and its confidence interval is 
below and to the left of the lines in the rightmost graphs 
of Figure~\ref{fig:CIvsParam_Lattice_Graph} we can say (with 95\% confidence) 
that ($kl$) and ($e$) are complements. 
For economies whose value of $\sigma$ is below and to the right of the line, 
we can say (with 95\% confidence) that ($kl$) and ($e$) are substitutes. 
For economies whose value of $\sigma$ is above those lines, 
we cannot say with 95\% confidence whether ($kl$) and ($e$) are substitues or complements.

The right column of graphs in Figure~\ref{fig:CIvsParam_Lattice_Graph} shows 
$CI_{95\%,\sigma}$ vs. $\sigma$ for the three quantifications of energy 
(thermal energy, $q$, exergy, $x$, and useful work, $u$). 
We note that energy ($e$) unambiguously complements 
capital stock and labor ($kl$) for most economies.

<<CIvsParam_Lattice_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="95\\% Confidence interval widths vs. parameter values for key energy parameters ($\\gamma$, $\\delta$, and $\\sigma$) of the Cobb-Douglas and CES models with energy. If an economy appears in the region above a dashed line, the 95\\% confidence interval for the parameter extends to or beyond $\\gamma \\le$ 0 (cannot say with 95\\% confidence that energy is important for the Cobb-Douglas model), $\\delta \\ge$ 1 (cannot say with 95\\% confidence that energy is important for the CES model), or $\\sigma$ = 1 (cannot say with 95\\% confidence whether ($kl$) and ($e$) are complements or substitutes in the CES model).">>=
CIvsParamLattice(countryAbbrevScaling=0.6)
@

Regardless of model (Cobb-Douglas or CES) or energy quantification 
($q$, $x$, or $u$), there are few economies for which we can say 
with high confidence (at the 95\% level) that energy plays 
a significant role in economic growth. 
However, in seeming contradiction, results for the CES model indicate 
that energy complements capital stock and labor, 
indicating that energy ($e$) is essential for capital stock ($k$) and 
labor ($l$) to generate economic growth. 
One possible explanation for this apparent contradiction is 
to consider energy as embedded within capital stock and labor: 
it takes energy to make both captial stock and labor available to generate economic growth.

**** Martin: Need additional interpretation here. ****

%++++++++++++++++++++++++++++++
\subsection{LINEX Models}
\label{sec:LINEX_Parameter_Estimation}
%++++++++++++++++++++++++++++++

The LINEX model is given in Equation~\ref{eq:LINEX}. 
Parameters $a_0$ and $c_t$ were estimated using the \texttt{nls} function 
in \texttt{R} \citep{R}.

Figures~\ref{fig:LINEXq_Resample_Graph}--\ref{fig:LINEXu_Resample_Graph}
show the coefficients estimated from resampled data.

<<LINEXq_Resample_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="LINEX model (with $q$) resampling results.">>=
# linexResamplePlot(energyType="Q")
linexResamplePlot(energyType="Q")
@

<<LINEXx_Resample_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="LINEX model (with $x$) resampling results.">>=
linexResamplePlot(energyType="X")
@

<<LINEXu_Resample_Graph, fig.pos="H", fig.width=ninePanelGraphWidth, fig.height=threePanelGraphHeight, fig.cap="LINEX model (with $u$) resampling results.">>=
linexResamplePlot(energyType="U")
@

Figure~\ref{fig:LINEXq_Params_Graph} shows values and 95\% confidence intervals 
for the parameters for the LINEX model (with $q$).

>>>>

\section*{Notes from Martin}

@@

1. Set scene through energy economic modelling papers. 
Point out that PF approaches are alive and well (Stern, Tol, Gerlagh, World Bank, IMF), 
short lit review as part of a longer introduction.  
Mention competing PF approaches and parameter choice. 
On what basis selection and how estimate parameters?

2. Point out the need for adequate statistical testing and economic interpretation, 
citing some earlier critical papers on PF limitations. Take issue with

3. Spell out the contribution we will be making in this paper, 
we are taking the approach that all models are wrong but some are useful. 
We are after generating insights even after deeper assumptions of econ growth theory are very limiting.

4. Introduce coordinates of analysis.

The remainder of this paper is organized as follows: **** complete this sentence ****.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parameter Estimation}
\label{sec:ParameterEstimation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section should cover both technique and results. 
Organize along coordinates of analysis.

1. Discuss parameter estimation for each model here.

2. Show scatter plot results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion of Results}
\label{sec:DiscussionOfResults}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Drive towards new insights from our stats tested comparative approach
